\begin{abstract}
Previous research suggests media attention may cause support for populist right-wing parties, but extant evidence remains arguable and mostly limited to proportional representation systems in which such an effect would be most likely. At the same time, in the United Kingdom's first-past-the-post system, an ongoing political and regulatory debate revolves around whether the media give disproportionate coverage to the populist right-wing UK Independence Party (UKIP). Thus, we use a mixed-methods approach to investigate the causal dynamics of UKIP support and media coverage as an especially valuable case. Vector autoregression (VAR) using monthly, aggregate time-series data from January 2004 to September 2015 provides new evidence consistent with a model in which media coverage drives party support, but party support does not drive media coverage. Additionally, qualitative investigation of the dynamics suggests that in at least two key periods of stagnating or declining support for UKIP, media coverage increased and was followed by increases in public support. Overall the findings show that media coverage can and does appear to drive public support in a substantively important fashion irreducible to previous levels of public support, even in a national institutional environment least supportive of such an effect. The findings have direct and troubling implications for contemporary political and regulatory debates in the United Kingdom and potentially liberal democracies more generally.\footnote{Justin Murphy (\href{http://jmrphy.net}{jmrphy.net}, \href{http://twitter.com/jmrphy}{@jmrphy}) is Assistant Professor of Politics at the University of Southampton. Daniel Devine is a post-graduate researcher at the University of Southampton. This manuscript was prepared for the 2016 annual meeting of the Political Studies Association in Brighton, UK. Comments, questions, and suggestions are welcome and can be emailed to \href{mailto:j.murphy@soton.ac.uk}{j.murphy@soton.ac.uk}. If you wish to cite, please use the citation and DOI provided with the most up-to-date version of this article, which will be maintained at \href{http://j.mp/ukip-media}{j.mp/ukip-media}.}
\end{abstract}
\doublespacing
\setlength\parindent{24pt}