---
title: "Does Public Support for UKIP Drive Their Media Coverage or Does Media Coverage Drive Support for UKIP?"
author:
- Justin Murphy
- Daniel Devine
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: mystyles.sty
      before_body: prefix.tex
    keep_tex: yes
    latex_engine: xelatex
    toc: no
  html_document:
    toc: no
fontsize: 12pt
geometry: margin=1in
linkcolor: blue
classoption: article
bibliography: mybib.bib
csl: jepop.csl
---

# Introduction

If the visibility of a political party in the media shapes the public support it receives, then the degree to which the media gives attention to different political parties can have significant implications for democracy. In the United Kingdom, critics allege that the media pays disproportionate attention to the populist, right-wing UK Independence Party (UKIP) but media elites claim that media coverage of UKIP is driven by increasing public support for the party. Descriptively, media attention to UKIP is greater than that given to other, similarly sized parties on the right as well as the left [@Goodwin:03LtHfhh; @Stevenson:wo; @Soussi:vy], but UK media regulator Ofcom as well as the BBC have publicly defended the attention paid to UKIP on grounds of public support for the party [@Sweeney:wp; @Wintour:vf]. Implied in this elite reasoning is a causal model, namely that public support drives media coverage rather than vice-versa.

Yet previous research from proportional representation systems suggests that public support does not drive media coverage for populist right-wing parties, but rather media coverage drives their public support [@Boomgaarden:2007ia; @Boomgaarden:2009ke; @vliegenthart_anti-immigrant_2012]. By leveraging this insight to investigate the causal dynamics of UKIP support and media coverage, we fill an important gap in current research on the visibility-support nexus and contribute pragmatically relevant insights to a contentious public policy debate of broad social significance [@Gerring:2015ub]. First, we contribute to current research on the visibility-support nexus by testing a key insight from this research in a new institutional context where the hypothesized relationship should be less likely. Because proportional representation systems are associated with a greater number of small parties [@Duverger:1972wk] and they tend to produce more diverse news [@Benson:2009kb; @Sheafer:2009hi; @Kumlin:2001iq; @Stromback:2006ht; @Baum:2012je], research confined to such systems is arguably most likely to reflect a model in which media coverage generates support for populist right-wing parties. In a first-past-the-post system, where we typically expect only two parties and media to be less diverse, these institutional pressures make it more difficult for the media to generate support for smaller populist, right-wing parties. Thus, testing this theory with time-series data from a first-past-the-post system contributes to either refining the scope conditions of previous research (in the case of unexpected findings) or else extending and strengthening our confidence in the media-support relationship. Secondly, we contribute to a pressing regulatory question in UK national politics, as the democratic quality of UK media regulation with respect to political party favouritism, especially regarding populist right-wing parties, remains on public trial. This article lends insight into the causal dynamics implied but rarely if ever tested within such popular policy debates.

The article begins by outlining the theory before moving to a discussion of our data, method and research strategy. We then present quantitative and qualitative analyses of the relationship between UKIP support and UKIP media coverage. A final section concludes.

# Theory

A large body of research suggests that mass media coverage, as the primary channel through which the electorate receives information about politicians and parties, affects many different aspects of electoral politics [@norris_virtuous_2000; @paletz_political_1996; @beck_social_2002; @dalton_partisan_1998]. If media coverage of political parties is driven by public support for the parties--even if media coverage then increases public support further--it could be argued that media are facilitating popular sovereignty. On the other hand, if media coverage independently changes public support rather than reflects it, this would represent a point of crucial possible distortion in the functioning of a democracy. The latent normative motivation for the present investigation is whether the quantity of UKIP's media coverage represents a form of media bias which generates rather than reflects public opinion, or if the media's fascination with UKIP is a democratically appropriate effect of public opinion. 

One current of previous research on the dynamics of media coverage and party support finds evidence consistent with the argument that the quantity of media coverage given to a political party drives public support for that party. Walgrave and De Swert [-@walgrave_making_2004] find that, in time-series data from Belgium, the evidence reflects a model in which newspapers and television stations helped to increase the electoral results of the Vlaams Blok by emphasising political issues owned by the extreme right-wing party.  Boomgaarden and Vliegenthart [-@Boomgaarden:2007ia; @vliegenthart_why_2010] find that in the Netherlands, quantity of media coverage on immigration-related topics is associated with a subsequent increase in the vote-share for anti-immigrant parties, controlling for objective factors such as levels of immigration. Boomgaarden and Vliegenthart [-@Boomgaarden:2009ke] also find, using time-series from Germany, that media coverage of immigrant actors is associated with subsequent change in public attitudes toward immigration, conditional on objective factors such as immigration levels. While much of the previous research above considers the political implications of issue coverage in the media, Vliegenhart, Boomgaarden, and Van Spanje [-@vliegenthart_anti-immigrant_2012] advance this current further by analyzing time-series on the coverage of parties and public support for anti-immigrant parties per se in Belgium, Netherlands, and Germany. That study finds evidence suggesting that party and party leader visibility is associated with the electoral outcomes of the parties, but not vice-versa. In another study, media coverage was found to be one of the best predictors of electoral success in the 2007 Dutch election [@hopmann_effects_2010]. Finally, it has been shown that in the Netherlands, media coverage of Fortuyn appears to have improved polling performance of the party before the 2002 election [@koopmans_rise_2009].

Considering research at the individual level, panel data from the Netherlands suggests that media coverage drives perceptions of right-wing populist politicians as well as mainstream politicians [@Bos:2011iy]. Media coverage has also been found to help explain individual-level party preferences in Germany [@semetko_germanys_1994] and the Netherlands [@oegema_personalization_2009]. Based on this previous research, we test the following hypothesis.

H1: *Increases in media coverage lead to increased public support, controlling for previous levels of public support.*

It is also theoretically plausible, as some scholars have argued, that changes in party support lead to changes in media coverage [@pauwels_reassessing_2010]. As Vliegenthart and Boomgaarden [-@vliegenthart_why_2010] consider, quantity of media coverage may be driven by the power and position of political figures. This pattern has been observed, in some cases, in America [@sellers_winning_2007] and Switzerland [@tresch_politicians_2009]. Sellers [-@sellers_winning_2007] finds that the types of events U.S. Senators hold, and the guests of those events, affects the number of news stories written. Tresch [-@tresch_politicians_2009] finds that the amount of coverage given to Swiss legislators is most importantly a function of leadership and authority criteria related to the individual politicians. Although both of these studies focus on politicians rather than political parties per se, they suggest that variable aspects of political entities have predictable effects on media visibility. In a study on the diffusion of populist discourse in the media, Rooduijn [-@rooduijn_mesmerising_2014] argues from a study of five Western European countries (Italy, France, Germany, Netherlands, and United Kingdom) the electoral success of populist parties affects the degree of populism in the media.\footnote{Interestingly, in the study by Rooduijn, UKIP is classified as the least successful case of a populist party, based on their electoral results as of 2005, yet populism in British newspapers in 2005 is near that found in Netherlands and Germany and greater than that found in France. Although the findings are interpreted as electoral politics driving media content, Rooduijn's data show that in the UK at least, populism in the media was comparatively high in cross-national perspective before UKIP rose to its recent prominence.} There has been surprisingly little scholarship in this field of research in relation to either the UK or UKIP. As a rare example, Deacon and Wring [-@deacon_uk_2015] offer a case study of newspaper coverage of UKIP over a similar time period covered in this article. They conclude that when media coverage did increase, this was because UKIP's political standing made them hard to ignore. Therein, they offer a causal logic that it was the political support which drove media coverage rather than the reverse.

In line with this current of research, British media and media regulators have publicly argued media coverage given to political parties is based on public support for the parties. In its draft electoral guidelines published in January 2015, the BBC classified UKIP as deserving a degree of coverage comparable to the “larger parties," because they "demonstrated a substantial increase in electoral support,” as measured by electoral and polling results, between 2010 and 2015 [@Sweeney:wp; @BBC:R6UMvIKM]. Ofcom, the UK broadcast regulator, also included UKIP as a "major party" for the purposes of the 2015 General Election and local elections in England and Wales [@Ofcom:2015tt], also explicitly on the grounds of improving electoral and polling results since 2010 [@Wintour:vf]. Based on this current of previous research and the stated reasoning of elite entities with uniquely strong influence on media agendas, we propose the following additional hypothesis opposite to H1.

H2: *Increases in public support for UKIP lead to increased media coverage, controlling for previous levels of media coverage.*

The remainder of the paper sets out to investigate these two hypotheses. The following section discusses the data and method we pursue before we then present our findings.

# Data, Method, and Research Strategy

To measure public support for UKIP, we gathered monthly aggregate polling data on vote intentions from Ipsos MORI [@IpsosMORI:gm2fXYNK]. Specifically, we constructed the variable *Support* from the percentage of respondents reporting an intention to vote for UKIP according to the Ipsos MORI polling for each month. For most months, this was straightforward because the Ipsos MORI poll is approximately monthly. For months with multiple polls, we used the poll closest to the middle of the month.\footnote{A drawback of this choice is that some polling information is lost, as some polls were not integrated into the dataset. An alternative would be to average all the polls for each month, but this would lead each monthly average to reflect different parts of each month (for instance, if one month has two polls only in the first half, and another month has two polls only in the second half). Because our main interest relates to dynamics, it seems more important to have consistent measures reflecting as close as possible the middle of each month, at the cost of some information loss, than to include more polls but inconsistently reflect different parts of each month.} For the very few months with no poll or a poll at the border between the previous or following month, the value was counted as missing and then all missing values were linearly interpolated. To measure media coverage of UKIP, we gathered monthly counts of all UK national newspaper reports mentioning either "UKIP" or "UK Independence Party" from the database Nexis.\footnote{Duplicate articles defined by Nexis's definition of high similarity were excluded.} This resulted in 65,416 articles over the time period covered. There have been criticisms of such computer-assisted approaches, mostly notably by Althaus et al [-@althaus_using_2001], but we follow Boomgaarden and Vliegenthart [-@Boomgaarden:2007ia] in believing that, for these types of study, this is a reasonable and valuable way of measuring media coverage. This is the most efficient way to analyse large amounts of media content over a long period of time, an approach which is especially suitable for our present purposes given that we are only looking at quantity or intensity of converage (i.e., the number of articles each month). 

\setlength\parindent{0pt}

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=2.7, fig.width=7}
require(lubridate)
require(ggplot2)
require(timeSeries)
require(tseries)
require(reshape2)
require(scales)
require(vars)
require(car)
require(dynlm)
require(stargazer)
require(sandwich)
require(forecast)
require(zoo)
require(xts)

options(scipen=999)
options(digits=2)
set.seed(345)
options(xtable.comment = FALSE)

#setwd("C:/Users/Daniel/Dropbox/ukip_media/article")
setwd("~/Dropbox/gh_projects/ukip_media/article")
#df<-read.csv("C:/Users/Daniel/Dropbox/ukip_media/data/ukip_media.csv")
df<-read.csv("~/Dropbox/gh_projects/ukip_media/data/ukip_media.csv")
df$Date<-mdy(df$Date)

df$UKIP.Vote<-interpNA(df$UKIP.Vote, method = "linear")

unscaled<-df

modelvars<-df[c("Date", "UKIP.Vote", "UKIP.Articles", "General.Election2005", "General.Election2010",
                "General.Election2015", "Euro.Election2004","Euro.Election2009", "Euro.Election2014")]

cbbPalette <- c("#D55E00", "#0072B2","#009E73","#999999", "#56B4E9","#CC79A7")

ggplot(df, aes(x=Date, y=UKIP.Vote)) +
  geom_line(colour="#0072B2", linetype=2) + theme_bw() + labs(y="Vote Intention (%)", x="") +
  theme(axis.text.y = element_text(size=12.45)) +
  labs(title="Dynamics of UKIP Support and Media Coverage")
```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=2.6, fig.width=7}
ggplot(df, aes(x=Date, y=UKIP.Articles)) +
  geom_line(colour="#D55E00", linetype=1) + theme_bw() + labs(y="Articles", x="") +
  theme(axis.text.y = element_text(size=7.25))
```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=2.6, fig.width=7, fig.cap="Dynamics of UKIP Support and Media Coverage"}

df$UKIP.Articles<-scale(df$UKIP.Articles)[,]
df$UKIP.Vote<-scale(df$UKIP.Vote)[,]

# Both series in one plot

graph.df<-subset(df, select=c("Date", "UKIP.Articles", "UKIP.Vote"))
molten<-melt(graph.df, id.vars=c("Date"))
ggplot(molten, aes(x=as.Date(Date))) +
    geom_line(aes(y=value, colour=variable, linetype=variable)) +
    scale_colour_manual(values=cbbPalette) +
  scale_x_date(labels = date_format("%Y"),
#               breaks = "6 months",
               limits = as.Date(c('2004-02-01','2015-09-01'))) +
  theme_bw() +
    theme(legend.position="bottom",
          legend.title=element_blank()) +
#  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(axis.text.y = element_text(size=12.45)) +
  labs(x="Date", y="Standardized Values") +
  theme(legend.position="none")
```

\setlength\parindent{24pt}

The variable *Articles* reflects the number of articles Nexis returns from the first day of each month until the last day of each month. Figure 1 provides a summary view of the two main variables of interest. The dotted line represents *Support* and the solid line represents *Articles*. Raw values are displayed in the first two (top) panels. For ease of direct comparison the bottom panel displays standardized scores in which each value is derived by subtracting the mean of the particular time-series and dividing by one standard devation.

It is also plausible that elections have an independent effect on coverage and support due to general increased media attention and campaigning. For this reason, we have included eponymous dummy variables for the months of each national and European election within the sampling period. The elections included are three European elections (June 2004, June 2009 and May 2014) and three general elections (May 2005, May 2010, and May 2015). European elections coincide with local elections in the UK.

In the present analysis we do not consider public opinion on particular political issues, measures of objective political or policy dynamics, or the visibility of party leaders in the media, for several reasons. The first and main reason is dictated by our problem-driven approach. Because our contribution to the literature is motivated by a particular debate in the politics of British media, we focus on the parameters of that debate, which have revolved around party coverage. Although UKIP's controversial leader Nigel Farage is likely a significant aspect of UKIP's media visibility, coverage of Farage is almost certainly highly correlated with coverage of the party, as Vliegenhart, Boomgaarden, and Van Spanje find of party and leader coverage in multiple other Western European countries. Second, Vliegenhart, Boomgaarden, and Van Spanje also find that media coverage of parties is, overall, more relevant than party leader as a predictor of party support [@vliegenthart_anti-immigrant_2012, 333]. While it is possible that phenomena such as objective immigration levels, media coverage of immigration, and/or public opinion on immigration may affect both UKIP party coverage and public support for UKIP, it is not theoretically straightforward that they should affect one of our main variables more, or sooner, than the other. Because we lack any particular theoretical perspective on such possibilities, and there are many additional causal factors which could arguably be included in this system, we refrain from proliferating additional variables [@Achen:2006fp].

We first use econometric techniques to test for, and distinguish the ordering of, potential causal dynamics between media coverage and public support for UKIP. An ideal approach to testing the presented hypotheses is vector autoregression (VAR) with Granger causality tests [@brandt2007multiple; @vliegenthart_anti-immigrant_2012]. Specifically, we estimate a VAR by OLS per equation, using the following form:

\begin{equation}
 \label{eq:VAR}
    y_t = A_1 y_{t-1} + … + A_p y_{t-p} + D_t + u_t
\end{equation}

where $y_t$ is a $K \times 1$ vector of endogenous variables and $u_t$ is the error term. In our case the endogenous variables are \emph{Support} and \emph{Articles}. The coefficient matrices $A_1$, …, $A_p$ are of dimension $K \times K$. By convention $p$ denotes the "order" of the VAR, or the number of lags used. Typically this is determined empirically, as we do below. In addition, $D_t$ refers to a vector of exogenous regressors. In our case the exogenous regressors include a constant term, a trend term, the dummy variable for UK General Election months, and the dummy variable of European election months. We then use the conventional F-type Granger-causality test for each of the two endogenous variables in the system. The vector of endogenous variables $y_t$ is divided into two vectors $y_1t$ and $y_2t$ of dimensionality ($K_1 \times 1$) and ($K_2 \times 1$) with $K = K_1 + K_2$ [@Pfaff:2008wk]. The null hypothesis is that no lags of variable $y_{1t}$ are significant in the equation for variable ${y}_{2t}$. If $\alpha_{21, i} = 0$ for $i = 1$, $2$, …, $p$, we say that $y_{1t}$ does not "Granger-cause" $y_{2t}$.


Additionally, a brief qualitative historical analysis of the dynamics is conducted to further probe any potential causal process(es). It is arguably a general blindspot of quantitative time-series research to neglect inquiry into the subsantive historical processes corresponding to the statistical properties of time-series data. In particular, the substantive nature of the puzzle at hand requires the identification of a historical narrative which would not necessarily follow from a statistical fact such as Granger causality. Even with econometric evidence suggesting an association in one direction or the other, it would remain unclear whether the historical unfolding of such dynamics may imply a substantively significant issue for the core democratic function under consideration.

For instance, it could be the case that, formally, media coverage Granger-causes public support *and* that exogenous increases in media coverage played no particularly important role in the rise of UKIP support. This is because statistical properties of time-series in no way preclude the fact that the historically key moments of UKIP's rise could have been random or contingent consequences of other factors. Additionally, it is always possible in any particular historical process that $Y_1$ has an average effect on $Y_2$ which is statistically significant but in key, contingent moments certain shifts in $Y_2$ may explain unique changes in $Y_1$ in a fashion which is not statistically distinguishable. In the latter case, media-caused increases in public support might themselves be responding to, and amplifying, contingent but exogenous increases in public support in an arguably democracy-consistent fashion, even if increases in support do not statistically predict increases in media coverage.

To provide the strongest possible investigation of the role media has played in the rise of UKIP support, we will need to assess the degree to which increases in media coverage were followed by increases in public support for UKIP following *stagnant or decreasing* levels of support in preceding months. We will then also need to assess the degree to which such identifiable historical moments were related to the relatively few key moments in which support for UKIP rises most dramatically. We explore these substantive questions with a brief but detailed narrative of the political events and media themes which lie behind our time-series data.

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, fig.height=4, fig.width=8}
# adf.test(df$UKIP.Articles)  # non-stationary
# adf.test(df$UKIP.Vote) # non-stationary

modelvars$UKIP.Articles.d<-diff(zoo(scale(modelvars$UKIP.Articles)), na.pad=T)
modelvars$UKIP.Articles.d.l<-lag(modelvars$UKIP.Articles.d, k=-1, na.pad=T)
modelvars$UKIP.Articles.d<-as.numeric(modelvars$UKIP.Articles.d)
modelvars$UKIP.Articles.d.l<-as.numeric(modelvars$UKIP.Articles.d.l)
modelvars$UKIP.Vote.d<-diff(zoo(scale(modelvars$UKIP.Vote)), na.pad=T)
modelvars$UKIP.Vote.d.l<-lag(modelvars$UKIP.Vote.d,k=-1, na.pad=T)
modelvars$UKIP.Vote.d<-as.numeric(modelvars$UKIP.Vote.d)
modelvars$UKIP.Vote.d.l<-as.numeric(modelvars$UKIP.Vote.d.l)

modelvars$UKIP.Articles.notlog<-scale(modelvars$UKIP.Articles)
modelvars$UKIP.Vote.notlog<-scale(modelvars$UKIP.Vote)

modelvars$UKIP.Articles<-log(modelvars$UKIP.Articles)
modelvars$UKIP.Vote<-log(modelvars$UKIP.Vote)

modelvars$General.Elections<-ifelse(modelvars$General.Election2005==1 | modelvars$General.Election2010==1 |
                                     modelvars$General.Election2015==1, 1, 0)
modelvars$EU.Elections<-ifelse(modelvars$Euro.Election2004==1 | 
                                modelvars$Euro.Election2009==1 |
                                modelvars$Euro.Election2014==1,
                              1, 0)

modelvars$UKIP.Articles<-scale(modelvars$UKIP.Articles)
modelvars$UKIP.Vote<-scale(modelvars$UKIP.Vote)

diffvars<-data.frame()  # difference variables and lag dates/elections to match
diffvars<-as.data.frame(cbind(modelvars$Date[2:length(modelvars$Date)],
                              diff(modelvars$UKIP.Vote),
                              diff(modelvars$UKIP.Articles),
                              modelvars[2:length(modelvars$Date), 4:11]))
names(diffvars)<-c("Date", "UKIP.Vote", "UKIP.Articles", "General.Election2005", "General.Election2010",
                    "General.Election2015", "Euro.Election2004","Euro.Election2009", "Euro.Election2014",
                   "General.Elections", "EU.Elections")
row.names(diffvars)<-NULL


# adf.test(diffvars$UKIP.Articles)  # stationary after differencing
# adf.test(diffvars$UKIP.Vote)   # stationary after differencing

# VARselect(diffvars[2:3], type="both", exogen=diffvars[10:11])$selection[1]

varmodel<-VAR(diffvars[c(3,2)], lag.max=10, ic="AIC", type="both", exogen=diffvars[10:11])
# serial.test(varmodel)  # big p = good
# arch.test(varmodel)    # big p = good
# stab<-stability(varmodel)
# plot(stab)
# summary(varmodel)
# summary(varmodel, equation="UKIP.Articles")
# summary(varmodel, equation="UKIP.Vote")
```

# Findings and Discussion

Because both variables are non-stationary, vector autoregression is estimated with first differences of each variable. Optimal lag length is determined by the Aikeke Information Criterion to be VAR(3). The model includes a constant and a trend term. Diagnostics suggest that using the log of each variable before differencing reduces heteroskedasticity and serial correlation of errors. The models displayed here all pass the standard ARCH-LM and Portmanteau tests for non-constant error variance and serial correlation of errors, respectively. Finally, diagnostics show no evidence of significant temporal instability (see Supplementary Information).

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, results='asis'}

stargazer(varmodel$varresult$UKIP.Vote, varmodel$varresult$UKIP.Articles,
          title="Vector Autoregression",
          column.labels=c("$\\Delta Support$", "$\\Delta Articles$"),
          dep.var.labels=c(""),
          covariate.labels=c("$\\Delta Articles_{t-1}$",
                             "$\\Delta Support_{t-1}$",
                             "$\\Delta Articles_{t-2}$",
                             "$\\Delta Support_{t-2}$",
                             "$\\Delta Articles_{t-3}$",
                             "$\\Delta Support_{t-3}$",
                             "Constant",
                             "Trend",
                             "General Elections",
                             "EU Elections"),
          header=FALSE)

```

Initial VAR results show little evidence that changes in public support predict media coverage, but statistically significant evidence that media coverage drives public support. As the numerical results and the Impulse Response plots show, there is no statistically discernable correlation between past changes in public support and changes in media coverage, whereas past changes in media coverage have a statistically significant correlation with future changes in public support. As reported in Table 2, Granger causality tests support this interpretation, with the latter relationship identified in the model very unlikely to be observed by chance alone (p<.05).

We note there are limitations of the data which may make it difficult to identify the full range of causal effects in a VAR approach. First, it is possible that monthly measures are too infrequent to capture causal effects if the real lag between effects is shorter than one month. Also, importantly, structural tests on all models suggest strong evidence of instantaneous causality. Thus, the VAR results suggest clear but imperfect and, for reasons discussed above, inherently limited evidence for Hypothesis 1 that increases in media coverage lead to increases in public support. The VAR results provide no evidence for Hypothesis 2, that increases in public support lead to increases in media coverage. Given the problem of instantaneous causality, we cannot rule out the possibility that both variables drive each other in periods shorter than one month or that both variables are driven by some third unobserved variable. Nonetheless, the stylized empirical fact of our data is that media coverage Granger-causes public support and not vice-versa.

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, results='asis'}

gc.articles<-causality(varmodel, cause="UKIP.Vote", vcov.=vcovHC(varmodel))
gc.vote<-causality(varmodel, cause="UKIP.Articles", vcov.=vcovHC(varmodel))

gc.articles.result<-as.data.frame(unlist(gc.articles$Granger))
gc.vote.result<-as.data.frame(unlist(gc.vote$Granger))

results<-merge(gc.vote.result, gc.articles.result, by="row.names")

results<-results[3:6,1:3]
results$Row.names<-c("P-value", "DF1", "DF2", "F-test")
names(results)<-c("", "Support", "Articles")
results$Support<-as.numeric(as.character(results$Support))
results$Articles<-as.numeric(as.character(results$Articles))

stargazer(results,
          summary=F,
          title="Granger Causality Tests",
          style="APSR",
          digits=3,
          rownames=F,
          header=F)

```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, fig.height=4, fig.width=8, fig.cap='Impulse Response Plot Shows Effect on Articles from an Exogenous Increase in Support'}
set.seed(666)


# irf(varmodel, impulse="UKIP.Vote", response="UKIP.Articles", cumulative=F, boot=F)

plot(irf(varmodel, impulse="UKIP.Vote", response="UKIP.Articles", cumulative=F, boot=T,
         seed=345, n.ahead=12, ortho=T), 
     main="Impulse Response of Articles from Support")
```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, fig.height=4, fig.width=8, fig.cap='Impulse Response Plot Shows Effect on Support from an Exogenous Increase in Articles'}
set.seed(666)
plot(irf(varmodel, impulse="UKIP.Articles", response="UKIP.Vote", cumulative=F, boot=T,
         seed=345, n.ahead=24, ortho=T),
     main="Impulse Response of Support from Articles")

# fanchart(predict(varmodel, n.ahead=3, dumvar=diffvars[1:3,10:11]))

```

# Qualitative Analysis

To what extent are the statistical regularities identified by the vector autoregression historically significant causal factors in the rise of public support for UKIP? To facilitate a qualitative investigation of the dynamics, we quantitatively identifed months which meet criteria similar to the concept of Granger causality. Any month ($t$) that is immediately preceded by two months ($t-1$, $t-2$) of stagnating or declining public support but increased media coverage, we designate as a month of "uncaused" media increase or media "bias" for short. Symmetrically, any month that is immediately preceded by two months of stagnating or declining media coverage but increasing public support, we consider a month of "uncaused" or exogenously increasing public support. To mitigate the probability we will be counting mere noise as meaningful increases, we count as increases only those greater than .05 standard deviations and all other months as "stagnating or decreasing." Figure 2 presents the standardized values of each time series with dashed and green vertical lines indicating months of uncaused media 'bias' and grey, dotted vertical lines indicating months of uncaused increases in public support. Even a first, superficial consideration of Figure 2 reveals that increases in media coverage unwarranted by public support are not only roughly as frequent as uncaused increases in public support, but they are found at multiple pivotal months in periods of the most dramatic increases in UKIP's public support. To be clear, we are not claiming to pinpoint key moments of causal effect; in any particular point of the time-series, it is impossible to know whether a pattern is random noise or a true "signal" of one variable causing a change in another variable. Rather, we take the evidence from the VAR to be our warrant for exploring the qualitative data in search of possible examples whereby the substantive significance of the statistical evidence may either be better illustrated or possibly discounted due to untheorized contingencies. Based on our preliminary explorations summarized in Figure 2, we focus especially on two key periods: from July to September of 2012, and the second half of 2013.

UKIP, formed in 1993, began fielding European parliamentary candidates in 1994 and British parliamentary candidates in 1997. Since then, the party has enjoyed mixed but notable increases in public support and in electoral outcomes, particularly in the European parliament where the party was the largest in the 2014 election. Until the 2015 general election, UKIP’s domestic electoral success had been much less impressive, receiving just 3.1\% of the vote in 2010. Like other small or new parties, it has a history of infighting, changes of direction and leadership, and problems with financial mismanagement [@Whitaker:2011gi]. As recently as 2011, a lack of media attention was cited as a factor in UKIP’s poor performance, as well as credibility and relatively few activists [@ford_strategic_2012]. Indeed, the historical pattern of both media coverage and public support for UKIP over much of its recent history, from 2004 to 2009, was a series of small increases which consistently returned to low baseline quantities of little political consequences [@murphy_ukip_2015].

The party experienced its first bump in both coverage and voting intention in 2004 with the European election, in which they received 16\% of the vote, where coverage reached 829 articles in a single month, their record amount of coverage at the time and the greatest amount of coverage the party would experience until 2012. During this spike, both media coverage and voting intention increase proportionately and as would be expected if coverage was driven by public opinion: Figure 2 indicates no bias or exogenous increases of support in this instance. Following this, both coverage and support decay and return to poltically negligible levels. Over the next eight years, there are a range of events that do no attract very much media attention or public support; indeed, events occur between these years that are similar to those that will occur in later years but they fail to generate the extraordinary media attention gained by such events in later years. The vast majority of coverage is "in passing," such as everday reports of election results, or else it is negative, regarding claims of fraud and infighting. Indeed, Figure 2 shows that this period was characterised by exogenous increases of support not tracked by media coverage, supporting the claim that a lack of media coverage failed to facilitate public support through this period [@ford_strategic_2012]. 

Apart from the 2005 election, in which UKIP received little coverage and performed poorly [@_election_2005; @morris_election_2005], UKIP saw little change in public support or media coverage until the European elections of 2009. There is a small boost in both support and coverage in April 2006, when David Cameron calls the party ‘fruitcakes, loonies’ and ‘closet racists’ [@white_ukip_2006]. Interestingly, this rise in media coverage was followed by a small but sustained boost in public support, which persisted for three months.  In April 2008, Conservative MP Bob Spink defected, giving UKIP their first MP which generated very little coverage, despite being called a coup [@winnett_tory_2008]. Even the European election in 2009, in which UKIP came in second place, generated far less coverage than the 2005 European election, where the party came in third place. Despite this, it was still hailed as a ‘political earthquake’ [@watt_european_2009] and garnered coverage for UKIP’s leader Nigel Farage.

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=4, fig.width=7, fig.cap="Standardized Time-Series, Green Lines Indicate Media \"Bias\" and Grey Lines Indicate Exogenous Increases in Support"}

qual.graph.df<-subset(modelvars, select=c("Date", "UKIP.Articles.d", "UKIP.Vote.d", "UKIP.Vote.d.l", 
                                          "UKIP.Articles.d.l", "UKIP.Vote.notlog", "UKIP.Articles.notlog"))
qual.graph.df$UKIP.Articles.notlog<-as.numeric(qual.graph.df$UKIP.Articles.notlog)
qual.graph.df$UKIP.Vote.notlog<-as.numeric(qual.graph.df$UKIP.Vote.notlog)
qual.graph.df<-qual.graph.df[complete.cases(qual.graph.df),]

qual.graph.df$bias<-ifelse(qual.graph.df$UKIP.Articles.d>.05 &
                             qual.graph.df$UKIP.Vote.d<.05 &
                             qual.graph.df$UKIP.Vote.d.l<.05, 1, 0)

qual.graph.df$exog<-ifelse(qual.graph.df$UKIP.Articles.d<.05 &
                             qual.graph.df$UKIP.Articles.d.l<.05 &
                             qual.graph.df$UKIP.Vote.d>.05, 1, 0)

##### Master

qual.graph.df$Support<-qual.graph.df$UKIP.Vote.notlog
qual.graph.df$Articles<-qual.graph.df$UKIP.Articles.notlog
molten<-melt(qual.graph.df[c("Articles","Support", "Date")], id.vars=c("Date"))

ggplot(molten, aes(x=as.Date(Date))) +
    geom_line(aes(y=value, colour=variable, linetype=variable)) +
    scale_colour_manual(values=cbbPalette) +
  scale_x_date(labels = date_format("%m/%Y"),
               breaks = pretty_breaks(),
               limits = as.Date(c('2004-01-01','2015-09-01'))) +
  theme_bw() +
    theme(legend.position="bottom",
          legend.title=element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(axis.text.y = element_text(size=12.45)) +
  labs(x="Date", y="Standardized Values") +
  geom_vline(xintercept = as.numeric(as.Date(qual.graph.df$Date[qual.graph.df$bias==1])), linetype = "dashed", color = "green") +
    geom_vline(xintercept = as.numeric(as.Date(qual.graph.df$Date[qual.graph.df$exog==1])), linetype = "dotted", color = "grey")

```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=4, fig.width=4, fig.cap="Standardized Time-Series, April 2012 to January 2014"}

ggplot(molten, aes(x=as.Date(Date))) +
    geom_line(aes(y=value, colour=variable, linetype=variable)) +
    scale_colour_manual(values=cbbPalette) +
  scale_x_date(labels = date_format("%m/%Y"),
               breaks = pretty_breaks(10),
               limits = as.Date(c('2012-05-01','2014-01-01'))) +
  theme_bw() +
    theme(legend.position="bottom",
          legend.title=element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(axis.text.y = element_text(size=12.45)) +
  labs(x="Date", y="Standardized Values") +
  geom_vline(xintercept = as.numeric(as.Date(qual.graph.df$Date[qual.graph.df$bias==1])), linetype = "dashed", color = "green") +
    geom_vline(xintercept = as.numeric(as.Date(qual.graph.df$Date[qual.graph.df$exog==1])), linetype = "dashed", color = "grey")
```


Following this, there are at least two occasions where media coverage both precedes and seems unrelated to UKIP's public support, which may then have generated further increases in popular support [@murphy_ukip_2015]. To aid our investigation, we refer to Figure 5, which shows Figure 4 at higher resolution for the months between May 2012 and January 2014. This period includes a by-election in Corby, followed by the UKIP party conference and a controversial Rotherham by-election [@wainwright_rotherham_2012]. In July 2012 UKIP's public support was unremarkably near its average and was declining from June, after it had been stagnant since May. But media coverage held relatively steady, slightly decreasing once but slightly increasing twice (and slightly increasing overall) from June to September. It is only at this point that public support increases notably from September to October and is followed by a spike in media coverage that likely represented a moment of positive feedback ending in the first truly significant rise of UKIP into mainstream public consciousness.

Between August and November 2012, the amount of articles covering UKIP increased from 198 to 948, the most they had ever received in one month at the time, beyond three standard deviations from their long-term mean. To be clear, this dramatic surge of UKIP support appears to launch with a moment of postive feedback betwen support and media coverage, genuinely containing a notable spike of public support at the beginning. However, the main quantitative and qualitative implication of this particular period, is that the months of July and September 2012 are months in which media coverage is slightly increasing despite stagnant or declining levels of public support, and it is these dynamically unresponsive months of media coverage that precede the spike in support observed in October. Of course, it is impossible to distinguish these slight increases in media coverage in July and September from random noise in the polling; but from the statistical analysis we have reason to believe such moments of unresponsively increasing media coverage are at least comparatively more likely to be predictve of changes in support than vice versa. Thus, while it would be impossible to demonstrate conclusively that these months of media coverage played a causal role in the dramatic rise of support achieved by November, our model suggests it is more likely these unresponsively stable and slightly increasing months of media coverage played a causal role in the increased support of October, than it is that the increased support in October played a causal role in the then-highest level of media coverage seen in November. In turn, the unprecedently high levels of media coverage in November likely played more of a role in the following spike of support than the October spike in support played in the November spike in media coverage. This interpretation is enhanced by the additional fact that after the spike in support of October, November returned to the lowest level of support observed in several months. Again, while we cannot confidently read causal dynamics in particular data points, the point is that the increase in support of October, which ostensibly seems to be followed by a spike in media coverage ultimately leading to UKIP's real debut, is a less plausible interpretation of the data than one based on Hypothesis 1.

Now consider the period between July 2013 and December 2014. Despite public support declining rapidly and steadily from its high point in April 2013, media coverage from July to August increases. Public support continues to decline through August until November. While media coverage appears to adjust dynamically downward after it's "uncaused" increase of August, yet again in November media coverage stabilizes and slightly increases. It is only at this point in November that support ends its long and steady decline and yet again begins another substantial increase until it returns back to the high levels of April 2013.  Again, in these two months we identify apparently minor but potentially crucial non-dynamically-responsive levels of media coverage which may be functioning as a floor preventing support from continuing to decline and making possible the surge beginning from November 2013. While of course these spikes and drops in support may just be volatility around UKIP's new, higher mean levels of support, the key point here is only to explore and give possible instances of the statistical findings. Unlike the previous instance of "bias" explored above, where political events such as by-elections and the party conference season may have played roles, in this case there are no obvious and directly party-related events shaping the dynamics in this period. However, one key event which may have played a role at this time is the lifting of work restrictions on Romanian and Bulgarian nationals which occurred in January 2014 [@martin_immigration_2013], with media coverage intensifying in the months leading up to January. The increased salience of issues related to migration and the European Union may help to explain changes in media coverage independent of UKIP's support. Interestingly, considerable coverage also surrounded Farage's comment, in December 2013, that Britain should accept Syrian refugees [@goodman_does_2013].

Previous studies have relied on statistical models similar to the one we have presented here. However, this may ignore interesting dynamics hidden within the data about what is happening in the relationship between media coverage and public opinion. A qualitative appreciation of the data indicates at least two key examples where increased media coverage unwarranted by changes in public support take place in key periods of UKIP's rise.

# Conclusion

This study has made three contributions. Firstly, to our knowledge this is one of the first articles to study the dynamics of right-wing populist party support and quantity of media coverage in the context of a majoritarian system and the UK in particular; previous research has primarily focused on other Western European democracies such as Belgium, the Netherlands and Germany. Despite the change in political system, our findings support those of [@vliegenthart_why_2010; @vliegenthart_anti-immigrant_2012], as we find quantitative and qualitative evidence that media coverage has played a unique causal role in increasing support for UKIP, in a fashion irreducible to previous levels of support or election outcomes.

Secondly, this article contributes to currently on-going efforts to advance the methodological aspects of research on media and public opinion [@Vliegenthart:2014di]. Unlike many quantitative studies, we provide an analytically sophisticated qualitative investigation of our statistical findings. Most previous research on the visibility-support nexus relies primarily on statistical evidence, which cannot necessarily address important questions relating to the substantive historical narrative of a particular political party. We find that, in two periods, increases in media coverage came after two months of stagnating or declining public support but was then followed by historically pivotal increases in support. While we cannot claim these periods are definitive instances of causality, they show that the particular and contingent historical unfolding of UKIP is consistent with the inference, suggested by our statistical analysis, that media coverage played a unique and important causal role in the rise of public support for UKIP.

Perhaps most importantly, these findings are of significance to contemporary public debate in the UK concerning the perception that unfair quantities of media coverage are given to UKIP. Some have argued that extensive media coverage of UKIP is justified due to public support for the party. The findings here, on the other hand, suggest this is an unacceptable argument: the extraordinary media coverage which has been given to UKIP cannot be explained or defended on grounds of public support. We find that media coverage has no reliable relationship to public support in the one month, two months, or three months before a particular month of coverage. Indeed, we find that coverage may have independently and uniquely driven some of the very public support which media regulators would later point to as their justification for the extraordinary coverage given to UKIP. Our findings therefore raise serious questions for the function of media coverage in a democratic political system, because they suggest that unelected and unrepresentative actors (the media) may be systematically shaping public opinion toward, and the fortunes of, certain political parties in contradiction to organic levels of public support for those parties.

As with all studies, there are limitations to the present study and certain areas for future research may be indicated. We have left aside the question of leader effects, given previously ambiguous findings. We do not undertake any form of content analysis to address the actual content of the coverage in question, but only look at the quantity of articles. It is possible that, by disaggregating the coverage further, different types of coverage may have different effects; it would also be interesting to see whether the positivity or negativity of coverage matters for future levels of public support. Similarly, we do not disaggregate between types of newspapers, such as broadsheet or tabloid, which may offer different types of coverage and have different effects. We also only focus on print media. This means that we have not accounted for the effect of visual and social media which may play a role in dynamics of party support.

\pagebreak

# Supplementary Information
```{r supplementary, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE}
#plot(varmodel, ask=FALSE)
stab<-stability(varmodel)
plot(stab)
```

\pagebreak

# References

\singlespacing

\raggedright

\setlength\parindent{0pt}