---
title: Does Public Support for UKIP Drive Media Coverage or Does Media Coverage Drive
  Support for UKIP?
author: "Justin Murphy and Daniel Devine"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: mystyles.sty
      before_body: prefix.tex
    keep_tex: yes
    latex_engine: xelatex
    toc: no
  html_document:
    toc: no
fontsize: 12pt
geometry: margin=1in
linkcolor: blue
classoption: article
bibliography: mybib.bib
csl: jepop.csl
---

# Introduction

If the visibility of a political party in the media shapes the public support it receives, then the degree to which the media gives attention to different political parties can have significant implications for democracy. In the United Kingdom, critics allege that the media pays disproportionate attention to the populist, right-wing UK Independence Party (UKIP) but media elites claim that media coverage of UKIP is driven by increasing public support for the party. Descriptively, media attention to UKIP is greater than that given to other, similarly sized parties on the right as well as the left [@Goodwin:03LtHfhh; @Stevenson:wo; @Soussi:vy], but UK media regulator Ofcom as well as the BBC have publicly defended the attention paid to UKIP on grounds of public support for the party [@Sweeney:wp; @Wintour:vf]. Implied in this elite reasoning is a causal model, namely that public support drives media coverage rather than vice-versa.

Yet previous research from proportional representation systems suggests that public support does not drive media coverage for populist right-wing parties, but rather their media coverage drives their public support [@Boomgaarden:2007ia; @Boomgaarden:2009ke; @vliegenthart_anti-immigrant_2012]. By leveraging this insight to investigate the causal dynamics of UKIP support and media coverage, we fill an important gap in current research on the visibility-support nexus and contribute pragmatically relevant insights to a contentious public policy debate of broad social significance [@Gerring:2015ub]. First, we contribute to current research on the visibility-support nexus by testing a key insight from this research in a new institutional context where hypothesized relationship should be less likely. Because proportional representation systems are associated with a greater number of small parties [@Duverger:1972wk] and they tend to produce more diverse news [@Benson:2009kb; @Sheafer:2009hi; @Kumlin:2001iq; @Stromback:2006ht; @Baum:2012je], research confined to such systems is arguably most likely to reflect a model in which media coverage generates support for populist right-wing parties. In a first-past-the-post system, where we typically expect only two parties and media to be less diverse, these institutional pressures make it more difficult for the media to generate support for smaller populist, right-wing parties. Thus, testing this theory with time-series data from a first-past-the-post system contributes to either refining the scope conditions of previous research (in the case of unexpected findings) or else extending and strengthening our confidence in the media-support relationship. Additionally, we contribute to a pressing regulatory question in UK national politics, as the democratic quality of UK media regulation with respect to political party favouritism, especially regarding populist right-wing parties, remains on public trial. This article lends insight into the causal dynamics implied but rarely if ever tested within such popular policy debates.

The article begins by outlining the theory from which we are generating our expectations, and outlines our hypotheses. It then moves on to a discussion on the data, method and research strategy we use to investigate these hypotheses, before discussing the findings and implications. We then present a qualitative analysis of the growth of UKIP to support the quantitative findings, before concluding. 

# Theory

A large body of research suggests that mass media coverage, as the primary channel through which the electorate receives information about politicians and parties, affects many different aspects of electoral politics [@norris_virtuous_2000; @paletz_political_1996; @beck_social_2002; @dalton_partisan_1998]. If media coverage of political parties is driven by public support for the parties--even if media coverage then increases public support further--it could be argued that media are facilitating popular sovereignty. On the other hand, if media coverage independently changes public support rather than reflects it, this would represent a point of crucial possible distortion in the functioning of a democracy. The latent normative motivation for the present investigation is whether the quantity of UKIP's media coverage represents a form of media bias which generates rather than reflects public opinion, or if the media's fascination with UKIP is a democratically appropriate effect of public opinion. 

One current of previous research on the dynamics of media coverage and party support finds evidence consistent with the argument that the quantity of media coverage given to a political party drives public support for that party. Walgrave and De Swert [-@walgrave_making_2004] find that, in time-series data from Belgium, the evidence reflects a model in which newspapers and television stations helped to increase the electoral results of the Vlaams Blok by emphasising political issues owned by the extreme right-wing party.  Boomgaarden and Vliegenthart [-@Boomgaarden:2007ia; @vliegenthart_why_2010] find that in the Netherlands, quantity of media coverage on immigration-related topics is associated with a subsequent increase in the vote-share for anti-immigrant parties, controlling for objective factors such as levels of immigration. Boomgaarden and Vliegenthart [-@Boomgaarden:2009ke] also find, using time-series from Germany, that media coverage of immigrant actors is associated with subsequent change in public attitudes toward immigration, conditional on objective factors such as immigration levels. While much of the previous research above considers the political implications of issue coverage in the media, Vliegenhart, Boomgaarden, and Van Spanje [-@vliegenthart_anti-immigrant_2012] advance this current further by analyzing time-series on the coverage of parties and public support for anti-immigrant parties per se in Belgium, Netherlands, and Germany. That study finds evidence suggesting that party and party leader visibility is associated with the electoral outcomes of the parties, but not vice-versa. In another study, media coverage was found to be one of the best predictors of electoral success in the 2007 Dutch election [@hopmann_effects_2010]. Finally, it has been shown that in the Netherlands, media coverage of Fortuyn appears to have improved polling performance of the party before the 2002 election [@koopmans_rise_2009].

Considering research at the individual level, panel data from the Netherlands suggests that media coverage drives perceptions of right-wing populist politicians as well as mainstream politicians [@Bos:2011iy]. Media coverage has also been found to help explain individual-level party preferences in Germany [@semetko_germanys_1994] and the Netherlands [@oegema_personalization_2009]. Based on this previous research, we test the following hypothesis.

H1: Increases in media coverage lead to increased public support, controlling for previous levels of public support and previous levels of media coverage. 

It is also theoretically plausible, as some scholars have argued, that changes in party support lead to changes in media coverage [@pauwels_reassessing_2010]. As Vliegenthart and Boomgaarden [-@vliegenthart_why_2010] consider, quantity of media coverage may be driven by the power and position of political figures. This pattern has been observed, in some cases, in America [@sellers_winning_2007] and Switzerland [@tresch_politicians_2009]. Sellers [-@sellers_winning_2007] finds that the types of events U.S. Senators hold, and the guests of those events, affects the number of news stories written. Tresch [-@tresch_politicians_2009] finds that the amount of coverage given to Swiss legislators is most importantly a function of leadership and authority criteria related to the individual politicians. Although both of these studies focus on politicians rather than political parties per se, they suggest that variable aspects of political entities have predictable effects on media visibility. In a study on the diffusion of populist discourse in the media, Rooduijn [-@rooduijn_mesmerising_2014] argues from a study of five Western European countries (Italy, France, Germany, Netherlands, and United Kingdom) the electoral success of populist parties affects the degree of populism in the media.\footnote{Interestingly, in the study by Rooduijn, UKIP is classified as the least successful case of a populist party, based on their electoral results as of 2005, yet populism in British newspapers in 2005 is near that found in Netherlands and Germany and greater than that found in France. Although the findings are interpreted as electoral politics driving media content, Rooduijn's data show that in the UK at least, populism in the media was comparatively high in cross-national perspective before UKIP rose to its recent prominence.}

In line with this current of research, British media and media regulators have publicly argued media coverage given to political parties is based on public support for the parties. In its draft electoral guidelines published in January 2015, the BBC classified UKIP as deserving a degree of coverage comparable to the “larger parties," because they "demonstrated a substantial increase in electoral support,” as measured by electoral and polling results, between 2010 and 2015 [@Sweeney:wp; @BBC:R6UMvIKM]. Ofcom, the UK broadcast regulator, also included UKIP as a "major party" for the purposes of the 2015 General Election and local elections in England and Wales [@Ofcom:2015tt], also explicitly on the grounds of improving electoral and polling results since 2010 [@Wintour:vf]. Based on this current of previous research and the stated reasoning of elite entities with uniquely strong influence on media agendas, we propose the following additional hypothesis opposite to H1.

H2: Increases in public support for UKIP lead to increased media coverage, controlling for previous levels of media coverage and previous levels of public support.

The following section discusses the data and method we pursue to test these two hypotheses.

# Data, Method, and Research Strategy

To measure public support for UKIP, we gathered monthly aggregate polling data on vote intentions from Ipsos MORI [@IpsosMORI:gm2fXYNK]. Specifically, we constructed the variable *Support* from the percentage of respondents reporting an intention to vote for UKIP according to the Ipsos MORI polling for each month. For most months, this was straightforward because the Ipsos MORI poll is approximately monthly. For months with multiple polls, we used the poll closest to the middle of the month.\footnote{A drawback of this choice is that some polling information is lost, as some polls were not integrated into the dataset. An alternative would be to average all the polls for each month, but this would lead each monthly average to reflect different parts of each month (for instance, if one month has two polls only in the first half, and another month has two polls only in the second half). Because our main interest relates to dynamics, it seems more important to have consistent measures reflecting as close as possible the middle of each month, at the cost of some information loss, than to include more polls but inconsistently reflect different parts of each month.} For the very few months with no poll or a poll at the border between the previous or following month, the value was counted as missing and then all missing values were linearly interpolated. To measure media coverage of UKIP, we gathered monthly counts of all UK national newspaper reports mentioning either "UKIP" or "UK Independence Party" from the database Nexis.\footnote{Duplicate articles defined by Nexis's definition of high similarity were excluded.} The variable *Articles* reflects the number of articles Nexis returns from the first day of each month until the last day of each month. Figure 1 provides a summary view of the two main variables of interest. The dotted line represents *Support* and the solid line represents *Articles*. Raw values are displayed in the first two (top) panels. For ease of direct comparison the bottom panel displays standardized scores in which each value is derived by subtracting the mean of the particular time-series and dividing by one standard devation.

It is also plausible that elections have an independent effect on coverage and support due to general increased media attention and campaigning. For this reason, we have included eponymous dummy variables for the months of each national and European election within the sampling period. The elections included are three European elections (June 2004, June 2009 and May 2014) and three general elections (May 2005, May 2010, and May 2015). European elections coincide with local elections in the UK.

In the present analysis we do not consider public opinion on particular political issues, measures of objective political or policy dynamics, or the visibility of party leaders in the media, for several reasons. The first and main reason is dictated by our problem-driven approach. Because our contribution to the literature is motivated by a particular debate in the politics of British media, we focus on the parameters of that debate, which have revolved around party coverage. Although UKIP's controversial leader Nigel Farage is likely a significant aspect of UKIP's media visibility, coverage of Farage is almost certainly highly correlated with coverage of the party, as Vliegenhart, Boomgaarden, and Van Spanje find of party and leader coverage in multiple other Western European countries. Second, Vliegenhart, Boomgaarden, and Van Spanje also find that media coverage of parties is, overall, more relevant than party leader as a predictor of party support [@vliegenthart_anti-immigrant_2012, 333]. While it is possible that phenomena such as objective immigration levels, media coverage of immigration, and/or public opinion on immigration may affect both UKIP party coverage and public support for UKIP, it is not theoretically straightforward that they should affect one of our main variables more, or sooner, than the other. Because we lack any particular theoretical perspective on such possibilities, and there are many additional causal factors which could arguably be included in this system, we refrain from proliferating additional variables [@Achen:2006fp].

We first use econometric techniques to test for, and distinguish the ordering of, potential causal dynamics between media coverage and public support for UKIP. An ideal approach to testing the presented hypotheses is vector autoregression (VAR) with Granger causality tests [@brandt2007multiple; @vliegenthart_anti-immigrant_2012]. Specifically, we estimate a VAR by OLS per equation, using the following form:

\begin{equation}
 \label{eq:VAR}
    y_t = A_1 y_{t-1} + … + A_p y_{t-p} + D_t + u_t
\end{equation}

where $y_t$ is a $K \times 1$ vector of endogenous variables and $u_t$ is the error term. In our case the endogenous variables are \emph{Support} and \emph{Articles}. The coefficient matrices $A_1$, …, $A_p$ are of dimension $K \times K$. By convention $p$ denotes the "order" of the VAR, or the number of lags used. Typically this is determined empirically, as we do below. In addition, $D_t$ refers to a vector of exogenous regressors. In our case the exogenous regressors include a constant term, a trend term, the dummy variable for UK General Election months, and the dummy variable of European election months. We then use the conventional F-type Granger-causality test for each of the two endogenous variables in the system. The vector of endogenous variables $y_t$ is divided into two vectors $y_1t$ and $y_2t$ of dimensionality ($K_1 \times 1$) and ($K_2 \times 1$) with $K = K_1 + K_2$ (CITE VAR PACKAGE). The null hypothesis is that no lags of variable $y_{1t}$ are significant in the equation for variable ${y}_{2t}$. If $α_{21, i} = 0$ for $i = 1$, $2$, …, $p$, we say that $y_{1t}$ does not "Granger-cause" $y_{2t}$.


Additionally, a brief qualitative historical analysis of the dynamics is conducted to further probe any potential causal process(es). In particular, the substantive nature of the puzzle at hand requires the identification of a causal narrative. Even with econometric evidence suggesting an association in one direction or the other, it would remain unclear whether the historical unfolding of such dynamics may imply a substantively significant issue for the core democratic function under consideration. In other words, we are not only interested in whether media coverage amplifies exogenous increases in support--this would be an important but not necessarily problematic finding from a democratic perspective. In particular we will have to assess the degree to which increases in media coverage could possibly generate support for UKIP *despite low, stagnant, or decreasing* levels of support, in a historically observable and non-trivial fashion. We explore this substantive question with a brief but detailed narrative of the real-world events which coincide behind the time-series plotted in Figure 1.


\pagebreak

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=2.5, fig.width=7}
require(lubridate)
require(ggplot2)
require(timeSeries)
require(tseries)
require(reshape2)
require(scales)
require(vars)
require(car)
require(dynlm)
require(stargazer)
require(sandwich)
require(forecast)
require(zoo)
require(xts)

#setwd("C:/Users/Daniel/Dropbox/ukip_media/article")
setwd("~/Dropbox/gh_projects/ukip_media/article")
#df<-read.csv("C:/Users/Daniel/Dropbox/ukip_media/data/ukip_media.csv")
df<-read.csv("~/Dropbox/gh_projects/ukip_media/data/ukip_media.csv")
df$Date<-mdy(df$Date)

df$UKIP.Vote<-interpNA(df$UKIP.Vote, method = "linear")

unscaled<-df

modelvars<-df[c("Date", "UKIP.Vote", "UKIP.Articles", "General.Election2005", "General.Election2010",
                "General.Election2015", "Euro.Election2004","Euro.Election2009", "Euro.Election2014")]

cbbPalette <- c("#D55E00", "#0072B2","#009E73","#999999", "#56B4E9","#CC79A7")

ggplot(df, aes(x=Date, y=UKIP.Vote)) +
  geom_line(colour="#0072B2", linetype=2) + theme_bw() + labs(y="Vote Intention (%)", x="") +
  theme(axis.text.y = element_text(size=12.45)) +
  labs(title="Dynamics of UKIP Support and Media Coverage")

ggplot(df, aes(x=Date, y=UKIP.Articles)) +
  geom_line(colour="#D55E00", linetype=1) + theme_bw() + labs(y="Articles", x="") +
  theme(axis.text.y = element_text(size=7.25))
```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=3, fig.width=7, fig.cap="Dynamics of UKIP Support and Media Coverage"}

df$UKIP.Articles<-scale(df$UKIP.Articles)[,]
df$UKIP.Vote<-scale(df$UKIP.Vote)[,]

# Both series in one plot

graph.df<-subset(df, select=c("Date", "UKIP.Articles", "UKIP.Vote"))
molten<-melt(graph.df, id.vars=c("Date"))
ggplot(molten, aes(x=as.Date(Date))) +
    geom_line(aes(y=value, colour=variable, linetype=variable)) +
    scale_colour_manual(values=cbbPalette) +
  scale_x_date(labels = date_format("%m/%Y"),
#               breaks = "6 months",
               limits = as.Date(c('2004-02-01','2015-09-01'))) +
  theme_bw() +
    theme(legend.position="bottom",
          legend.title=element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(axis.text.y = element_text(size=12.45)) +
  labs(x="Date", y="Standardized Values") +
  theme(legend.position="none")
```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, fig.height=4, fig.width=8}
# adf.test(df$UKIP.Articles)  # non-stationary
# adf.test(df$UKIP.Vote) # non-stationary

modelvars$UKIP.Articles.d<-diff(zoo(scale(modelvars$UKIP.Articles)), na.pad=T)
modelvars$UKIP.Articles.d.l<-lag(modelvars$UKIP.Articles.d, k=-1, na.pad=T)
modelvars$UKIP.Articles.d<-as.numeric(modelvars$UKIP.Articles.d)
modelvars$UKIP.Articles.d.l<-as.numeric(modelvars$UKIP.Articles.d.l)
modelvars$UKIP.Vote.d<-diff(zoo(scale(modelvars$UKIP.Vote)), na.pad=T)
modelvars$UKIP.Vote.d.l<-lag(modelvars$UKIP.Vote.d,k=-1, na.pad=T)
modelvars$UKIP.Vote.d<-as.numeric(modelvars$UKIP.Vote.d)
modelvars$UKIP.Vote.d.l<-as.numeric(modelvars$UKIP.Vote.d.l)

modelvars$UKIP.Articles.notlog<-scale(modelvars$UKIP.Articles)
modelvars$UKIP.Vote.notlog<-scale(modelvars$UKIP.Vote)

modelvars$UKIP.Articles<-log(modelvars$UKIP.Articles)
modelvars$UKIP.Vote<-log(modelvars$UKIP.Vote)

modelvars$General.Elections<-ifelse(modelvars$General.Election2005==1 | modelvars$General.Election2010==1 |
                                     modelvars$General.Election2015==1, 1, 0)
modelvars$EU.Elections<-ifelse(modelvars$Euro.Election2004==1 | 
                                modelvars$Euro.Election2009==1 |
                                modelvars$Euro.Election2014==1,
                              1, 0)

modelvars$UKIP.Articles<-scale(modelvars$UKIP.Articles)
modelvars$UKIP.Vote<-scale(modelvars$UKIP.Vote)

diffvars<-data.frame()  # difference variables and lag dates/elections to match
diffvars<-as.data.frame(cbind(modelvars$Date[2:length(modelvars$Date)],
                              diff(modelvars$UKIP.Vote),
                              diff(modelvars$UKIP.Articles),
                              modelvars[2:length(modelvars$Date), 4:11]))
names(diffvars)<-c("Date", "UKIP.Vote", "UKIP.Articles", "General.Election2005", "General.Election2010",
                    "General.Election2015", "Euro.Election2004","Euro.Election2009", "Euro.Election2014",
                   "General.Elections", "EU.Elections")
row.names(diffvars)<-NULL


# adf.test(diffvars$UKIP.Articles)  # stationary after differencing
# adf.test(diffvars$UKIP.Vote)   # stationary after differencing

# VARselect(diffvars[2:3], type="both", exogen=diffvars[10:11])$selection[1]

varmodel<-VAR(diffvars[c(3,2)], lag.max=10, ic="AIC", type="both", exogen=diffvars[10:11])
# serial.test(varmodel)  # big p = good
# arch.test(varmodel)    # big p = good
# stab<-stability(varmodel)
# plot(stab)
# summary(varmodel)
# summary(varmodel, equation="UKIP.Articles")
# summary(varmodel, equation="UKIP.Vote")
```

# Findings and Discussion

Because both variables are non-stationary, vector autoregression is estimated with first differences of each variable. Optimal lag length is determined by the Aikeke Information Criterion to be VAR(3). The model includes a constant and a trend term. Diagnostics suggest that using the log of each variable before differencing reduces heteroskedasticity and serial correlation of errors. Because VAR models have many paramaters to begin with, monthly indicators controlling for seasonality absorb crucial degrees of freedom and so are excluded in the intitial models but added in subsequent models. The models displayed here all pass the standard ARCH-LM and Portmanteau tests for non-constant error variance and serial correlation of errors, respectively. Finally, diagnostics show no evidence of significant temporal instability (see Supplementary Information).

Surprisingly, initial VAR results show little evidence that changes in public support predict media coverage, but significant evidence that media coverage drives public support. As the numerical results and the Impulse Response plots show, there is no statistically discernable correlation between past changes in public support and changes in media coverage, whereas past changes in media coverage have a statistically significant correlation with changes in public support. Granger causality tests support this interpretation, with only the latter relationship nearing conventional cutoffs of statistical significance (p<.08).

After including monthly indicators, however, the results reverse: while the coefficients reflecting the correlation between past changes in media coverage and public support do not change noticeably, they lose statistical significance, whereas the coefficients for the other model become significant and pass the test for Granger causality. Because the coefficients reflecting the correlation between past changes in media coverage and public support remain signed as predicted, the increased standard errors do not necessarily reflect the absence of a relationship but possibly only a lack of degrees of freedom due to the introduction of the seasonality indicators.

Additionally, there are limitations of the data which may make it difficult to identify causal effects in a VAR approach. First, it is possible that monthly measures are too infrequent to capture causal effects if the real lag between effects is more shorter than one month. Importantly, structural tests on all models suggest strong evidence of instantaneous causality.

Taken together, VAR results suggest qualified evidence for both directions of causality. While the results are sensitive to the specification, the results are consistent with the possibility that both variables drive each other, but that highly robust evidence of this in one model is not possible due to the nature of the data and the high-paramater demands of the VAR approach.
\pagebreak

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, results='asis'}


stargazer(varmodel$varresult$UKIP.Vote, varmodel$varresult$UKIP.Articles,
          column.labels=c("Vote", "Articles"),
          dep.var.labels=c(""),
          header=FALSE)


```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, results='asis'}

gc.articles<-causality(varmodel, cause="UKIP.Vote", vcov.=vcovHC(varmodel))
gc.vote<-causality(varmodel, cause="UKIP.Articles", vcov.=vcovHC(varmodel))

gc.articles.result<-as.data.frame(unlist(gc.articles$Granger))
gc.vote.result<-as.data.frame(unlist(gc.vote$Granger))

results<-merge(gc.vote.result, gc.articles.result, by="row.names")

results<-results[3:6,1:3]
results$Row.names<-c("P-value", "DF1", "DF2", "F-test")
names(results)<-c("", "Support", "Articles")
results$Support<-as.numeric(as.character(results$Support))
results$Articles<-as.numeric(as.character(results$Articles))

stargazer(results,
          summary=F,
          title="Granger Causality Tests",
          style="APSR",
          digits=3,
          rownames=F)

```


```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, fig.height=4, fig.width=8}
set.seed(666)


# irf(varmodel, impulse="UKIP.Vote", response="UKIP.Articles", cumulative=F, boot=F)

plot(irf(varmodel, impulse="UKIP.Vote", response="UKIP.Articles", cumulative=F, boot=T,
         seed=345, n.ahead=12, ortho=T), 
     main="Impulse Response of Articles from Support")

plot(irf(varmodel, impulse="UKIP.Articles", response="UKIP.Vote", cumulative=F, boot=T,
         seed=345, n.ahead=24, ortho=T),
     main="Impulse Response of Support from Articles")

# fanchart(predict(varmodel, n.ahead=3, dumvar=diffvars[1:3,10:11]))

```

# Qualitative Analysis

UKIP, formed in 1993, began fielding European parliamentary candidates in 1994 and British parliamentary candidates in 1997. Since then, the party has enjoyed increasing electoral success, particularly in the European parliament where the party was the largest in the 2014 election. Until the 2015 general election, UKIP’s domestic electoral success had been much less impressive, receiving just 3.1\% of the vote in 2010. Like other small or new parties, it has a history of infighting, changes of direction and leadership, and problems with financial mismanagement [@Whitaker:2011gi]. As recently as 2011, a lack of media attention was cited as a factor in UKIP’s poor performance, as well as credibility and few activists [@ford_strategic_2012]. Indeed, the historical pattern of both media coverage and voting intention has been one of short spikes followed by a usually drastic return to low baseline levels [@murphy_ukip_2015].

The party experienced its first bump in both coverage and voting intention in 2004 with the European election, in which they received 16\% of the vote, where coverage reached 829 articles in a single month, the highest to date and the highest the party would experience until 2012. During this spike, both media coverage and voting intention increase proportionately and as would be expected if coverage was driven by public opinion. Following this, the party experiences a slow decline in both coverage and support. Over the next eight years, there are a range of events which attract neither much media attention nor public support; indeed, similar events occur between these years to those that occur in later years but which generate much more media attention. The vast majority of coverage is either in passing, such as election results, or negative, such as claims of fraud and infighting. 

Apart from the 2005 election, in which UKIP received little coverage and performed very poorly [@_election_2005; @morris_election_2005], UKIP disappeared into the wilderness until the European elections of 2009. There is a small boost in both support and coverage in April 2006, when David Cameron calls the party ‘fruitcakes, loonies’ and ‘closet racists’ [@white_ukip_2006]. Interestingly, this rise in media coverage was followed by a small but sustained boost in public support, which persisted for three months.  In April 2008, a Conservative MP, Bob Spink, defected, giving UKIP their first MP which generated very little coverage, despite being called a coup [@winnett_tory_2008]. Even the European election in 2009, in which UKIP came second, generated far less coverage than the 2005 European election, where the party came third. Despite this, it was still hailed as a ‘political earthquake’ [@watt_european_2009] and garnered coverage for UKIP’s leader Nigel Farage.

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=4, fig.width=7, fig.cap="Standardized Time-Series, Green Lines Indicate Bias and Grey Lines Indicate Exogenous Increases in Support"}

qual.graph.df<-subset(modelvars, select=c("Date", "UKIP.Articles.d", "UKIP.Vote.d", "UKIP.Vote.d.l", 
                                          "UKIP.Articles.d.l", "UKIP.Vote.notlog", "UKIP.Articles.notlog"))
qual.graph.df$UKIP.Articles.notlog<-as.numeric(qual.graph.df$UKIP.Articles.notlog)
qual.graph.df$UKIP.Vote.notlog<-as.numeric(qual.graph.df$UKIP.Vote.notlog)
qual.graph.df<-qual.graph.df[complete.cases(qual.graph.df),]

qual.graph.df$bias<-ifelse(qual.graph.df$UKIP.Articles.d>.05 &
                             qual.graph.df$UKIP.Vote.d<.05 &
                             qual.graph.df$UKIP.Vote.d.l<.05, 1, 0)

qual.graph.df$exog<-ifelse(qual.graph.df$UKIP.Articles.d<.05 &
                             qual.graph.df$UKIP.Articles.d.l<.05 &
                             qual.graph.df$UKIP.Vote.d>.05, 1, 0)

##### Master

molten<-melt(qual.graph.df[c("UKIP.Vote.notlog", "UKIP.Articles.notlog", "Date")], id.vars=c("Date"))

ggplot(molten, aes(x=as.Date(Date))) +
    geom_line(aes(y=value, colour=variable, linetype=variable)) +
    scale_colour_manual(values=cbbPalette) +
  scale_x_date(labels = date_format("%m/%Y"),
               breaks = pretty_breaks(),
               limits = as.Date(c('2004-01-01','2015-09-01'))) +
  theme_bw() +
    theme(legend.position="bottom",
          legend.title=element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(axis.text.y = element_text(size=12.45)) +
  labs(x="Date", y="Standardized Values") +
  geom_vline(xintercept = as.numeric(as.Date(qual.graph.df$Date[qual.graph.df$bias==1])), linetype = "dashed", color = "green") +
    geom_vline(xintercept = as.numeric(as.Date(qual.graph.df$Date[qual.graph.df$exog==1])), linetype = "dashed", color = "grey")

```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=4, fig.width=7, fig.cap="Standardized Series May 2012 to January 2014"}

ggplot(molten, aes(x=as.Date(Date))) +
    geom_line(aes(y=value, colour=variable, linetype=variable)) +
    scale_colour_manual(values=cbbPalette) +
  scale_x_date(labels = date_format("%m/%Y"),
               breaks = pretty_breaks(10),
               limits = as.Date(c('2012-05-01','2014-01-01'))) +
  theme_bw() +
    theme(legend.position="bottom",
          legend.title=element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(axis.text.y = element_text(size=12.45)) +
  labs(x="Date", y="First Differences of Standardized Values") +
  geom_vline(xintercept = as.numeric(as.Date(qual.graph.df$Date[qual.graph.df$bias==1])), linetype = "dashed", color = "green") +
    geom_vline(xintercept = as.numeric(as.Date(qual.graph.df$Date[qual.graph.df$exog==1])), linetype = "dashed", color = "grey")
```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=4, fig.width=7, fig.cap="First Differences of Standardized Series, January 2013 to September 2014"}

ggplot(molten, aes(x=as.Date(Date))) +
    geom_line(aes(y=value, colour=variable, linetype=variable)) +
    scale_colour_manual(values=cbbPalette) +
  scale_x_date(labels = date_format("%m/%Y"),
               breaks = pretty_breaks(),
               limits = as.Date(c('2013-01-01','2014-09-01'))) +
  theme_bw() +
    theme(legend.position="bottom",
          legend.title=element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(axis.text.y = element_text(size=12.45)) +
  labs(x="Date", y="First Differences of Standardized Values") +
  geom_vline(xintercept = as.numeric(as.Date(qual.graph.df$Date[qual.graph.df$bias==1])), linetype = "dashed", color = "green") +
    geom_vline(xintercept = as.numeric(as.Date(qual.graph.df$Date[qual.graph.df$exog==1])), linetype = "dashed", color = "grey")
```

Following this, there are at least two occasions where media coverage both precludes and seems unrelated to UKIP's performance in opinion polls, which may then have generated further increases in popular support [@murphy_ukip_2015]. The first occurred between August and November 2012 when, at a point when UKIP's support continued to be unremarkable and followed a similar pattern as before, the amount of articles covering UKIP increased from 198 to 948 - the most they have ever received in one month. This process began with a by-election in Corby, followed by the UKIP party conference and a controversial Rotherham by-election [@wainwright_rotherham_2012]. The media coverage remains consistent despite unremarkable poll ratings until January 2013, when support for UKIP spikes and continues on an upward trajectory.

This boost in support continues until May 2013 - a time of local elections - before beginning to decline, slowly and consistently, from a high of 15\% in April to 8\% in November 2013: historically quite high, but not especially so. The second instance is from November 2013 to December 2014. Despite poll ratings being at their lowest for the entire year, media coverage begins to increase. It is only following this perhaps unwarranted coverage that UKIP's poll ratings begin to recover, and both continue on an upward trajectory through to the end of 2014. Unlike the previous increase in media coverage, where exogenous events such as by-elections and conference season were present, it is difficult to identify a particular incident which was behind the sudden increase in coverage. One potential mechanism was the lifting of work restrictions on Romanian and Bulgarian nationals [@martin_immigration_2013] which occurred in January 2014, with coverage intensifying in the months before. Whilst this was not directly related to UKIP, the increased salience of the issues of immigration and the European Union may have driven the media coverage independent of UKIP's support. Considerable coverage also surrounded Farage's comments in December 2013 that Britain should accept Syrian refugees [@goodman_does_2013]. This may have had the effect of not just generating coverage but also making UKIP appear a kinder or more acceptable alternative to the Conservatives. 

Previous studies have relied on statistical models similar to the one we have presented here. However, this may ignore interesting dynamics hidden within the data about what is actually happening in public discourse. A qualitative appreciation of the data indicates at least two recent examples where it appears current public support for UKIP is at least in part driven by past media coverage which in turn cannot be explained by public support. Whilst other events may have contributed to the first surge, it is hard to make the same case for the second surge.

# Conclusion

We have made three contributions with this study. Firstly, this is the first paper, in our knowledge, to address the visibility-support nexus in the context of the United Kingdom and a majoritarian system; previous research has primarily focused on other Western European democracies such as Belgium, the Netherlands and Germany. Despite the change in political system, our findings support those of [@vliegenthart_why_2010; @vliegenthart_anti-immigrant_2012], and find that the media can and have independently generated support for UKIP. We have left aside the question of leader effects, given previously ambiguous findings. There is also reason to believe that media dynamics are different in proportional systems, being more diverse in their coverage than in majoritarian systems. 

Secondly, we have contributed methodologically in two ways. We have offered qualitative evidence for our findings that, at least in this case, the media have generated support for a radical right-wing party. Previous research has offered only statistical evidence, which may not pick up questions relating to the historical narrative of the party in question. We address this gap here and find that the results are still robust. We have also contributed to a growing body of literature that uses time-series methods to address questions relating to the media [@Vliegenthart:2014di].

Perhaps most importantly, these findings are of significance to contemporary public debate in the UK concerning the role of the media and the perceived unfair coverage of UKIP. Some have argued that the media coverage of UKIP is justified due to its public support. The findings here, on the other hand, suggest that the causal arrow points the other way: that the media coverage drove the support of UKIP independent of its previous poll ratings. 

As with all studies, there are limitations and areas for future research. We do not undertake any form of content analysis to address the actual content of the coverage in question, but only look at the quantity of articles. It is possible that, by disaggregating the coverage further, different types of coverage change the findings; it would also be interesting to see whether how positive or negative the coverage is matters for changing public opinion. Similarly, we do not disaggregate between types of paper, such as broadsheet and tabloid, which offer different coverage and target a different readership. We also only focus on print media. This means that we have not accounted for the effect of visual and social media which may be contributing to this relationship. Despite these limitations, this paper provides a contribution to the continuing and growing debate concerning the media’s role in the growth of political parties and the wider ramifications for democratic debate. 


\pagebreak

```{r supplementary, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE}
#plot(varmodel, ask=FALSE)
stab<-stability(varmodel)
plot(stab)

```

\pagebreak

# References

\raggedright
