---
title: Does Public Support for UKIP Drive Media Coverage or Does Media Coverage Drive
  Support for UKIP?
author: "Justin Murphy"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: mystyles.sty
      before_body: prefix.tex
    keep_tex: yes
    latex_engine: xelatex
    toc: no
  html_document:
    toc: no
fontsize: 12pt
geometry: margin=1.25in
linkcolor: blue
classoption: article
bibliography: mybib.bib
csl: jepop.csl
---

# Introduction

If the visibility of a political party in the media shapes the public support it receives, then the degree to which the media gives attention to different political parties can have significant implications for democracy. In the United Kingdom, critics allege that the media pays disproportionate attention to the populist, right-wing UK Independence Party (UKIP) but media elites claim that media coverage of UKIP is driven by increasing public support for the party. Descriptively, media attention to UKIP is greater than that given to other, similarly sized parties on the right as well as the left [@Goodwin:03LtHfhh; @Stevenson:wo; @Soussi:vy], but UK media regulator Ofcom as well as the BBC have publicly defended the attention paid to UKIP on grounds of public support for the party [@Sweeney:wp; @Wintour:vf]. Implied in this elite reasoning is a causal model, namely that public support drives media coverage rather than vice-versa.

Yet previous research from proportional representation systems suggests that public support does not drive media coverage for populist right-wing parties, but rather their media coverage drives their public support [@Boomgaarden:2007ia; @Boomgaarden:2009ke; @vliegenthart_anti-immigrant_2012]. By leveraging this insight to examine the dynamics of UKIP support and media coverage, we fill an important gap in current research on the visibility-support nexus while directly addressing a pressing political research question (Dan: Can you find a citation(s) in a good UK politics journal calling more policy/impact-driven research agendas?) First, we contribute to current research on the visibility-support nexus by testing a key insight from this research in a new institutional context where hypothesized relationship should be less likely. Because proportional representation systems are associated with a greater number of small parties (Downs) and they tend to produce more diverse news [@Benson:2009kb; @Sheafer:2009hi; @Kumlin:2001iq; @Stromback:2006ht; @Baum:2012je], research confined to such systems is arguably most likely to reflect a model in which media coverage generates support for populist right-wing parties. In a first-past-the-post system where we typically expect only two parties and media to be less diverse, these institutional pressures make it more difficult for the media to generate support for smaller populist, right-wing parties. Thus, testing this theory with time-series data from a first-past-the-post system contributes to either refining the scope conditions of previous research (in the case of unexpected findings) or else extending and strengthening our confidence in the media-support dynamic. Additionally, we make contribute to a pressing regulatory question in UK national politics, as the democratic quality of UK media regulation with respect to political party favouritism, especially regarding populist right-wing parties, remains on public trial. This article lends insight into the causal dynamics usually implied in such political debates.

\pagebreak

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=2.5, fig.width=7}
require(lubridate)
require(ggplot2)
require(timeSeries)
require(tseries)
require(reshape2)
require(scales)
require(vars)
require(car)
require(dynlm)
require(stargazer)
require(sandwich)
require(forecast)

setwd("C:/Users/Daniel/Dropbox/ukip_media/article")
#setwd("~/Dropbox/gh_projects/ukip_media/article")
df<-read.csv("C:/Users/Daniel/Dropbox/ukip_media/data/ukip_media.csv")
#df<-read.csv("~/Dropbox/gh_projects/ukip_media/data/ukip_media.csv")
df$Date<-mdy(df$Date)

df$UKIP.Vote<-interpNA(df$UKIP.Vote, method = "linear")

unscaled<-df

modelvars<-df[c("Date", "UKIP.Vote", "UKIP.Articles", "General.Election2005", "General.Election2010",
                "General.Election2015", "Euro.Election2004","Euro.Election2009", "Euro.Election2014")]

cbbPalette <- c("#D55E00", "#0072B2","#009E73","#999999", "#56B4E9","#CC79A7")

ggplot(df, aes(x=Date, y=UKIP.Vote)) +
  geom_line(colour="#0072B2") + theme_bw() + labs(y="Vote Intention (%)", x="") +
  theme(axis.text.y = element_text(size=12.45)) +
  labs(title="Dynamics of UKIP Support and Media Coverage")

ggplot(df, aes(x=Date, y=UKIP.Articles)) +
  geom_line(colour="#D55E00") + theme_bw() + labs(y="Articles", x="") +
  theme(axis.text.y = element_text(size=7.25))
```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, fig.height=3, fig.width=7, fig.cap="Dynamics of UKIP Support and Media Coverage"}

df$UKIP.Articles<-scale(df$UKIP.Articles)[,]
df$UKIP.Vote<-scale(df$UKIP.Vote)[,]

# Both series in one plot

graph.df<-subset(df, select=c("Date", "UKIP.Articles", "UKIP.Vote"))
molten<-melt(graph.df, id.vars=c("Date"))
ggplot(molten, aes(x=as.Date(Date))) +
    geom_line(aes(y=value, colour=variable)) +
    scale_colour_manual(values=cbbPalette) +
  scale_x_date(labels = date_format("%m/%Y"),
               breaks = "6 months",
               limits = as.Date(c('2004-02-01','2015-09-01'))) +
  theme_bw() +
    theme(legend.position="bottom",
          legend.title=element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(axis.text.y = element_text(size=12.45)) +
  labs(x="Date", y="Standardized Values") +
  theme(legend.position="none")
```

# Theory

A large body of research suggests that mass media coverage, as the primary channel through which the electorate receives information about politicians and parties, affects many different aspects of electoral politics [@norris_virtuous_2000; @paletz_political_1996; @beck_social_2002; @dalton_partisan_1998]. If media coverage of political parties is driven by public support for the parties--even if media coverage then increases public support further--it could be argued that media are facilitating popular sovereignty. On the other hand, if media coverage independently changes public support rather than reflects it, this would represent a point of crucial possible distortion in the functioning of a democracy. The latent normative motivation for the present investigation is whether the quantity of UKIP's media coverage represents a form of media bias which produces rather than reflects public opinion, or if the media's fascination with UKIP is a democratically appropriate response to a rising dimension of public opinion. 

Previous research on the dynamics of media coverage and party support is consistent with the argument that quantity of media coverage drives party support in Belgium [@walgrave_making_2004], the Netherlands, and Germany [@Boomgaarden:2007ia; @Boomgaarden:2009ke; @vliegenthart_why_2010; @vliegenthart_anti-immigrant_2012]. At the individual level, panel data from the Netherlands suggests that media coverage drives perceptions of right-wing populist politicians as well as mainstream politicians [@Bos:2011iy]. At the level of electoral results, [@hopmann_effects_2010] showed that media coverage was one of the best predictors of electoral success in the 2007 Dutch election. [@koopmans_rise_2009] showed that in the Netherlands, media coverage of Fortuyn improved polling performance before the 2002 election. Media coverage has also been found to help explain party preferences in Germany [@semetko_germanys_1994] and the Netherlands [@oegema_personalization_2009].

Other scholars have addressed the question indirectly. In his study on the diffusion of the populist message in the media, [@rooduijn_mesmerising_2014] hypothesises that the electoral success of populist parties affects the degree and acceptability of populism in the media. Other authors in the literature explore the theoretical connection between media coverage and the rise of populist and right-wing parties, but offer little or no empirical evidence for the claim [@art_reacting_2007; @mudde_three_2013]. Nevertheless, there is a considerable body of literature which posits particular mechanisms which may offer a causal explanation for how the quantity of media coverage can increase support for a political party. The primary explanation revolves around issue saliency and the aligning of a party or a party leader with those salient issues [@brug_media_2006; @cushion_interpreting_2015; @dennison_immigration_2015]. In the case of UKIP, the party was strongly aligned with the issue of immigration and the European Union, and the increasing prominence of these issues in the media may have driven both coverage and support for the party. [@walgrave_making_2004] offer a similar analysis of the Vlaams Blok in Belgium, concluding that the media are at least part responsible for the growth of the right-wing party. Although that study does conclude in support of this relationship, they do not commit themselves to arguing a causal relationship.Thus we test the hypothesis that:

is based on data from proportional representation (PR) electoral systems.

although some have called these findings into question on the basis of methodological concerns [@pauwels_reassessing_2010].

In part, this may be because the application of time-series techniques to aggregate media data remains relatively under-explored [@Vliegenthart:2014di].

H1: Increases in media coverage lead to increased public support, controlling for previous levels of public support. 


It is also theoretically plausible that changes in public support for party receives may lead to changes in media coverage. As [@vliegenthart_why_2010] consider, this could be related to the power and position of political figures, referencing studies from both America [@sellers_winning_2007] and Switzerland [@tresch_politicians_2009] which highlight how the standing of a political actor influences media attention. Another possibility is that media coverage depends on the dynamics of the party itself [@pauwels_reassessing_2010]: in our case, the relative acceptability of UKIP's agenda as opposed to other British populist parties such as the British National Party may have contributed to the rise in coverage, as well as the popularity and charisma of Nigel Farage. Finally, political polling and the reporting of political polls is ubiquitous in British media, including tabloid papers running polls of their own readers. It could be the case that increasing media coverage is simply reflective of their standing in polls, or that there is a positive feedback mechanism operating between media coverage and polling. We therefore test this possibility, hypothesising that:

H2: Increases in public support for UKIP lead to increased media coverage, controlling for previous levels of media coverage

# Data, Method, and Research Strategy

To measure public support for UKIP, we gathered monthly aggregate polling data on vote intentions from Ipsos MORI [@IpsosMORI:gm2fXYNK]. Specifically, we constructed a variable from the percentage reporting an intention to vote for UKIP according to the Ipsos MORI poll closest to the middle of each month. For most months, this was straightforward because the Ipsos MORI poll is approximately monthly. For months with multiple polls, we used the poll closest to the middle of the month. For the very few months with no poll or a poll at the border between the previous or following month, the value was counted as missing and then all missing values were linearly interpolated. To measure media coverage of UKIP, I gathered monthly counts of all UK national newspaper reports mentioning either "UKIP" or "UK Independence Party" from the database Nexis.\footnote{Duplicate articles defined by Nexis's definition of high similarity were excluded.}

There is also reason to believe that elections themselves have an independent effect on coverage and support due to general increased media attention and through the running of public campaigns. For this reason, we have included dummy variables for each national and European election covered by the data. The elections included are three European elections (2004, 2009 and 2014) and three general elections (2005, 2010 and 2015). Usefully, the European elections coincide with local elections in the UK. 

Econometric techniques are used to test for, and distinguish the ordering of, potential causal dynamics between media coverage and public support for UKIP. A brief qualitative historical analysis of these dynamics will be used to better understand a potential causal process. In particular, the substantive nature of the puzzle at hand requires the identification of a causal narrative. Even with econometric evidence suggesting an independent causal effect from either one to the other, it would not be clear whether the historical unfolding of these causal dynamics implies a problem for democracy. We are not only interested in whether media coverage amplifies exogenous increases in support--this would be an important but not necessarily problematic finding from a democratic perspective--but whether increases in media coverage have generated support for UKIP despite low, stagnant, or decreasing levels of support.

# Analysis

## VAR

Because both variables are non-stationary, vector autoregression is estimated with first differences of each variable. Optimal lag length is determined by the Aikeke Information Criterion to be to be VAR(3). The model includes a constant and a trend term. Diagnostics suggest that using the log of each variable before differencing reduces heteroskedasticity and serial correlation of errors. Because VAR models have many paramaters to begin with, monthly indicators controlling for seasonality absorb crucial degrees of freedom and so are excluded in the intitial models but added in subsequent models. The models displayed here all pass the standard ARCH-LM and Portmanteau tests for non-constant error variance and serial correlation of errors, respectively. Finally, diagnostics show no evidence of significant temporal instability (see Supplementary Information).

Surprisingly, initial VAR results show little evidence that changes in public support predict media coverage, but significant evidence that media coverage drives public support. As the numerical results and the Impulse Response plots show, there is no statistically discernable correlation between past changes in public support and changes in media coverage, whereas past changes in media coverage have a statistically significant correlation with changes in public support. Granger causality tests support this interpretation, with only the latter relationship nearing conventional cutoffs of statistical significance (p<.08).

After including monthly indicators, however, the results reverse: while the coefficients reflecting the correlation between past changes in media coverage and public support do not change noticeably, they lose statistical significance, whereas the coefficients for the other model become significant and pass the test for Granger causality. Because the coefficients reflecting the correlation between past changes in media coverage and public support remain signed as predicted, the increased standard errors do not necessarily reflect the absence of a relationship but possibly only a lack of degrees of freedom due to the introduction of the seasonality indicators.

Additionally, there are limitations of the data which may make it difficult to identify causal effects in a VAR approach. First, it is possible that monthly measures are too infrequent to capture causal effects if the real lag between effects is more shorter than one month. Importantly, structural tests on all models suggest strong evidence of instantaneous causality.

Taken together, VAR results suggest qualified evidence for both directions of causality. While the results are sensitive to the specification, the results are consistent with the possibility that both variables drive each other, but that highly robust evidence of this in one model is not possible due to the nature of the data and the high-paramater demands of the VAR approach.
\pagebreak

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, fig.height=4, fig.width=8}
# adf.test(df$UKIP.Articles)  # non-stationary
# adf.test(df$UKIP.Vote) # non-stationary

modelvars$UKIP.Articles<-log(modelvars$UKIP.Articles)
modelvars$UKIP.Vote<-log(modelvars$UKIP.Vote)

modelvars$General.Elections<-ifelse(modelvars$General.Election2005==1 | modelvars$General.Election2010==1 |
                                     modelvars$General.Election2015==1, 1, 0)
modelvars$EU.Elections<-ifelse(modelvars$Euro.Election2004==1 | 
                                modelvars$Euro.Election2009==1 |
                                modelvars$Euro.Election2014==1,
                              1, 0)

modelvars$UKIP.Articles<-scale(modelvars$UKIP.Articles)
modelvars$UKIP.Vote<-scale(modelvars$UKIP.Vote)

diffvars<-data.frame()  # difference variables and lag dates/elections to match
diffvars<-as.data.frame(cbind(modelvars$Date[2:length(modelvars$Date)],
                              diff(modelvars$UKIP.Vote),
                              diff(modelvars$UKIP.Articles),
                              modelvars[2:length(modelvars$Date), 4:11]))
names(diffvars)<-c("Date", "UKIP.Vote", "UKIP.Articles", "General.Election2005", "General.Election2010",
                    "General.Election2015", "Euro.Election2004","Euro.Election2009", "Euro.Election2014",
                   "General.Elections", "EU.Elections")
row.names(diffvars)<-NULL


# adf.test(diffvars$UKIP.Articles)  # stationary after differencing
# adf.test(diffvars$UKIP.Vote)   # stationary after differencing

# VARselect(diffvars[2:3], type="both", exogen=diffvars[10:11])$selection[1]

varmodel<-VAR(diffvars[c(3,2)], lag.max=10, ic="AIC", type="both", exogen=diffvars[10:11])
# serial.test(varmodel)  # big p = good
# arch.test(varmodel)    # big p = good
# stab<-stability(varmodel)
# plot(stab)
# summary(varmodel)
# summary(varmodel, equation="UKIP.Articles")
# summary(varmodel, equation="UKIP.Vote")
```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, results='asis'}


stargazer(varmodel$varresult$UKIP.Vote, varmodel$varresult$UKIP.Articles)


```

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, results='asis'}

gc.vote<-causality(varmodel, cause="UKIP.Vote", vcov.=vcovHC(varmodel))
gc.articles<-causality(varmodel, cause="UKIP.Articles", vcov.=vcovHC(varmodel))

gc.vote.result<-as.data.frame(unlist(gc.vote$Granger))
gc.articles.result<-as.data.frame(unlist(gc.articles$Granger))

results<-merge(gc.vote.result, gc.articles.result, by="row.names")

results<-results[3:6,1:3]
results$Row.names<-c("P-value", "DF1", "DF2", "F-test")
names(results)<-c("", "Support", "Articles")
results$Support<-as.numeric(as.character(results$Support))
results$Articles<-as.numeric(as.character(results$Articles))

stargazer(results,
          summary=F,
          title="Granger Causality Tests",
          style="APSR",
          digits=3,
          rownames=F)

```


```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE, fig.height=4, fig.width=8}
set.seed(666)


# irf(varmodel, impulse="UKIP.Vote", response="UKIP.Articles", cumulative=F, boot=F)

plot(irf(varmodel, impulse="UKIP.Vote", response="UKIP.Articles", cumulative=F, boot=T,
         seed=345, n.ahead=12, ortho=T), 
     main="Impulse Response of Articles from Support")

plot(irf(varmodel, impulse="UKIP.Articles", response="UKIP.Vote", cumulative=F, boot=T,
         seed=345, n.ahead=24, ortho=T),
     main="Impulse Response of Support from Articles")

# fanchart(predict(varmodel, n.ahead=3, dumvar=diffvars[1:3,10:11]))

```

# Process-tracing

This paper has investigated a simple claim by the UK’s regulatory authorities and national broadcaster: that the increased media coverage of UKIP was justified by the party’s increased poll ratings. We have found that, on the contrary, on average media coverage preceded an increase in poll ratings controlling for previous media coverage, rather than media coverage increasing controlling for previous public support. (Justin: perhaps write some results up from the models?).


# Conclusion

We have made three contributions with this study. Firstly, this is the first paper, in our knowledge, to address the visibility-support nexus in the context of the United Kingdom and a majoritarian system; previous research has primarily focused on other Western European democracies such as Belgium, the Netherlands and Germany. Despite the change in political system, our findings support those of [@vliegenthart_why_2010; @vliegenthart_anti-immigrant_2012], and find that the media can and have independently generated support for UKIP. We have left aside the question of leader effects, given previously ambiguous findings. There is also reason to believe that media dynamics are different in proportional systems, being more diverse in their coverage than in majoritarian systems. 

Secondly, we have contributed methodologically in two ways. We have offered qualitative evidence for our findings that, at least in this case, the media have generated support for a radical right-wing party. Previous research has offered only statistical evidence, which may not pick up questions relating to the historical narrative of the party in question. We address this gap here and find that the results are still robust. We have also contributed to a growing body of literature that uses time-series methods to address questions relating to the media [@Vliegenthart:2014di].

Perhaps most importantly, these findings are of significance to contemporary public debate in the UK concerning the role of the media and the perceived unfair coverage of UKIP. Some have argued that the media coverage of UKIP is justified due to its public support. The findings here, on the other hand, suggest that the causal arrow points the other way: that the media coverage drove the support of UKIP independent of its previous poll ratings. 

As with all studies, there are limitations and areas for future research. We do not undertake any form of content analysis to address the actual content of the coverage in question, but only look at the quantity of articles. It is possible that, by disaggregating the coverage further, different types of coverage change the findings; it would also be interesting to see whether how positive or negative the coverage is matters for changing public opinion. Similarly, we do not disaggregate between types of paper, such as broadsheet and tabloid, which offer different coverage and target a different readership. We also only focus on print media. This means that we have not accounted for the effect of visual and social media which may be contributing to this relationship. Despite these limitations, this paper provides a contribution to the continuing and growing debate concerning the media’s role in the growth of political parties and the wider ramifications for democratic debate. 


\pagebreak

```{r supplementary, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, echo=FALSE}
#plot(varmodel, ask=FALSE)
stab<-stability(varmodel)
plot(stab)

```

\pagebreak



# References

\raggedright
